---
title: "Colossus limma DE by DFS risk group"
output: html_notebook
---

Gene-level differential expression (High vs Low DFS risk) for **Colossus transcriptomics** using limma.

**Risk groups:** `risk_grp` from `proteomics/survival_df_with_risk.rds`.

**Expression:** `colossus/rna.csv` (COL sample IDs mapped to G `patient_id` via master ID document).

**Data policy:** Originals copied to `analysis_working/colossus/`; outputs in `analysis_output/colossus/`.

**R environment:** Knit with `/work_space/envs/transcriptomics2/bin/R` (limma on `/work_space` disk).

```{r setup}
library(dplyr)
library(readr)
library(stringr)
if (requireNamespace("limma", quietly = TRUE)) library(limma)
```

```{r load-data}
if (file.exists("colossus/rna.csv")) {
  base_dir       <- "."
  proteomics_dir <- "../proteomics"
  integrated_dir <- "../Integrated "
} else if (file.exists("Transcriptomics/colossus/rna.csv")) {
  base_dir       <- "Transcriptomics"
  proteomics_dir <- "proteomics"
  integrated_dir <- "Integrated "
} else {
  stop("Knit from Transcriptomics/ or its parent directory.")
}

source(file.path(base_dir, "analysis_helpers.R"))
dirs <- setup_analysis_dirs(base_dir, "colossus")
output_dir <- dirs$output_dir

original_files <- c(
  expression       = file.path(base_dir, "colossus/rna.csv"),
  master IDs       = file.path(integrated_dir, "All_patinet_IDS_concatonated_masterdoc.csv"),
  survival w/ risk = file.path(proteomics_dir, "survival_df_with_risk.rds")
)
missing <- names(original_files)[!file.exists(original_files)]
if (length(missing) > 0) {
  stop(
    "Missing required original file(s):\n",
    paste0("  - ", missing, ": ", original_files[missing], collapse = "\n")
  )
}

working <- ensure_working_copies(original_files, dirs$work_dir)
survival_df <- readRDS(working["survival w/ risk"])
master_df <- read_csv(working["master IDs"], show_col_types = FALSE)
```

```{r build-matrix-and-meta}
expr_mat <- prepare_expr_samples_x_genes(
  read_csv(working["expression"], show_col_types = FALSE),
  "patient_id"
)
expr_mat <- filter_impute_expr(expr_mat)
cat("Genes x samples:", nrow(expr_mat), "x", ncol(expr_mat), "\n")

sample_ids <- colnames(expr_mat)
id_map <- build_colossus_id_map(master_df, sample_ids)

sample_meta <- id_map %>%
  rename(sample_id = sample) %>%
  left_join(
    survival_df %>% select(patient_id, risk_grp, rsf_risk),
    by = "patient_id"
  )

cat("Mapped to G patient ID:", sum(!is.na(sample_meta$patient_id)), "\n")
cat("Samples with Low/High risk:\n")
print(table(sample_meta$risk_grp, useNA = "always"))
```

```{r limma-de}
limma_out <- run_limma_high_vs_low(
  expr_mat     = expr_mat,
  sample_meta  = sample_meta,
  output_dir   = output_dir,
  file_prefix  = "Colossus"
)

cat("\nTop 5 up (High > Low):\n")
print(limma_out$top %>% filter(direction == "up"))

cat("\nTop 5 down (High < Low):\n")
print(limma_out$top %>% filter(direction == "down"))

sig <- limma_out$results %>% filter(adj.P.Val < 0.05)
cat("\nGenes FDR < 0.05:", nrow(sig), "\n")
if (nrow(sig) > 0) print(head(sig %>% select(gene, logFC, P.Value, adj.P.Val), 10))
```

```{r limma-figures}
export_limma_figures(
  results     = limma_out$results,
  expr_mat    = expr_mat,
  sample_meta = sample_meta,
  output_dir  = output_dir,
  file_prefix = "Colossus",
  cohort_title = "Colossus Transcriptomics"
)
```
