---
title: "Colossus xCell2 enrichment by DFS risk group"
output: html_notebook
---

Volcano differential analysis of xCell2 cell-type enrichment (High vs Low DFS risk) for the **Colossus transcriptomics** cohort.

**Data policy:** Original files are copied to `analysis_working/colossus/` at run time. All reads use those copies; outputs go to `analysis_output/colossus/`. Original datasets are never modified.

Colossus xCell sample IDs (`COL-xxx-R-xxx`) are mapped to **G patient IDs** via `All_patinet_IDS_concatonated_masterdoc.csv` before joining proteomics-derived risk groups.

**Mapping note:** Only `COL-001-R-*` samples currently link to G IDs in the master document. `COL-002-R-*` and `COL-268-R-*` samples are excluded from this risk-group analysis until mapping is extended.

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(EnhancedVolcano)
```

```{r load-data}
if (file.exists("collosus_transcriptomics_xcell_results.csv")) {
  base_dir       <- "."
  proteomics_dir <- "../proteomics"
  integrated_dir <- "../Integrated "
} else if (file.exists("Transcriptomics/collosus_transcriptomics_xcell_results.csv")) {
  base_dir       <- "Transcriptomics"
  proteomics_dir <- "proteomics"
  integrated_dir <- "Integrated "
} else {
  stop(
    "Cannot locate collosus_transcriptomics_xcell_results.csv. ",
    "Knit from Transcriptomics/ or its parent directory."
  )
}

source(file.path(base_dir, "analysis_helpers.R"))
dirs <- setup_analysis_dirs(base_dir, "colossus")

original_files <- c(
  xCell2 scores  = file.path(base_dir, "collosus_transcriptomics_xcell_results.csv"),
  survival w/risk = file.path(proteomics_dir, "survival_df_with_risk.rds"),
  master IDs     = file.path(integrated_dir, "All_patinet_IDS_concatonated_masterdoc.csv")
)
missing <- names(original_files)[!file.exists(original_files)]
if (length(missing) > 0) {
  stop(
    "Missing required original file(s):\n",
    paste0("  - ", missing, ": ", original_files[missing], collapse = "\n")
  )
}

working_files <- ensure_working_copies(original_files, dirs$work_dir)
xcell_res <- read_csv(working_files[[1]], show_col_types = FALSE)
survival_df <- readRDS(working_files[[2]])
master_df <- read_csv(working_files[[3]], show_col_types = FALSE)

sample_ids <- colnames(xcell_res)[-1]
cat("xCell samples:", length(sample_ids), "\n")
print(table(sub("-R-.*$", "", sample_ids)))
```

```{r build-id-map}
strip_fp <- function(x) str_replace(x, "-FP.*$", "")

master_col_map <- bind_rows(
  master_df %>%
    filter(!is.na(patient_id_g), patient_id_g != "") %>%
    transmute(patient_id = patient_id_g, col_id = strip_fp(COLOSSUS_ID)),
  master_df %>%
    filter(!is.na(patient_id_g), patient_id_g != "") %>%
    transmute(patient_id = patient_id_g, col_id = strip_fp(Old_COLOSSUS_ID)),
  master_df %>%
    filter(!is.na(patient_id_g), patient_id_g != "") %>%
    transmute(patient_id = patient_id_g, col_id = strip_fp(Sample_ID)),
  master_df %>%
    filter(!is.na(patient_id_g), patient_id_g != "") %>%
    transmute(patient_id = patient_id_g, col_id = strip_fp(Alternative_ID))
) %>%
  filter(!is.na(col_id), col_id != "", col_id != "NA") %>%
  distinct(col_id, .keep_all = TRUE)

id_map <- data.frame(sample = sample_ids, stringsAsFactors = FALSE) %>%
  left_join(master_col_map, by = c("sample" = "col_id"))

cat("Samples mapped to G patient ID:", sum(!is.na(id_map$patient_id)), "\n")
cat("Samples with proteomics risk:", sum(id_map$patient_id %in% survival_df$patient_id), "\n")

unmapped <- id_map %>% filter(is.na(patient_id))
if (nrow(unmapped) > 0) {
  cat("Unmapped prefixes:\n")
  print(table(sub("-R-.*$", "", unmapped$sample)))
}
```

```{r build-scores-long}
scores_long <- xcell_res %>%
  rename(cell_type = 1) %>%
  pivot_longer(-cell_type, names_to = "sample", values_to = "enrichment") %>%
  left_join(id_map, by = "sample") %>%
  left_join(
    survival_df %>% select(patient_id, risk_grp, rsf_risk),
    by = "patient_id"
  ) %>%
  filter(risk_grp %in% c("Low", "High"))

n_samples <- n_distinct(scores_long$sample)
n_low  <- sum(scores_long$risk_grp == "Low") / n_distinct(scores_long$cell_type)
n_high <- sum(scores_long$risk_grp == "High") / n_distinct(scores_long$cell_type)

cat("Analysis samples (Low/High risk):", n_samples, "\n")
cat("Low:", n_low, " High:", n_high, "\n")

differential_results <- scores_long %>%
  group_by(cell_type) %>%
  summarise(
    mean_high = mean(enrichment[risk_grp == "High"], na.rm = TRUE),
    mean_low  = mean(enrichment[risk_grp == "Low"], na.rm = TRUE),
    log2FC    = log2((mean_high + 1e-10) / (mean_low + 1e-10)),
    pval      = wilcox.test(enrichment ~ risk_grp)$p.value,
    .groups   = "drop"
  ) %>%
  mutate(neg_log10_p = -log10(pval))

differential_results_clean <- differential_results[
  is.finite(differential_results$log2FC) & !is.na(differential_results$pval),
]

print(differential_results_clean %>%
        filter(cell_type %in% c("B cell", "Exhausted CD8+ T cells")))
```

```{r volcano-export}
volcano_axes <- volcano_axis_limits(differential_results_clean)

p_volcano <- suppressWarnings(
  EnhancedVolcano(
    differential_results_clean,
    lab       = differential_results_clean$cell_type,
    x         = "log2FC",
    y         = "pval",
    pCutoff   = 0.05001,
    FCcutoff  = 0.5,
    title     = "Cell Type Enrichment: High vs Low Risk",
    subtitle  = paste0(
      "Colossus Transcriptomics | xCell2 scores | n = ",
      n_samples, " (Low ", n_low, ", High ", n_high, ")"
    ),
    xlab      = "Log2 Fold Change (High / Low)",
    ylab      = "-Log10(p-value)",
    pointSize = 2.0,
    labSize   = 3.5,
    drawConnectors        = TRUE,
    widthConnectors       = 0.4,
    maxoverlapsConnectors = Inf,
    xlim = volcano_axes$xlim,
    ylim = volcano_axes$ylim
  ) +
  theme(
    plot.margin     = margin(20, 60, 20, 60),
    legend.position = "bottom"
  )
)

print(p_volcano)

out_png <- file.path(dirs$output_dir, "Colossus_risk_enrichment_volcano.png")
ggsave(out_png, plot = p_volcano, width = 12, height = 8, dpi = 300, bg = "white")
cat("Saved:", out_png, "\n")

top_fc_table <- export_top_log2fc_table(
  differential_results,
  file.path(dirs$output_dir, "Colossus_top5_log2fc_up_down.csv")
)
print(top_fc_table)

differential_results_clean[
  order(abs(differential_results_clean$log2FC), decreasing = TRUE),
  c("cell_type", "log2FC", "pval")
]
```
