---
title: "Retrospective limma DE by DFS risk group"
output: html_notebook
---

Gene-level differential expression (High vs Low DFS risk) for **retrospective (Leuven) transcriptomics** using limma.

**Risk groups:** `risk_grp` from `proteomics/survival_df_with_risk.rds`.

**Expression:** `retrospective/Leuven_rlog_values.txt` (sample columns are G `patient_id`).

**Data policy:** Originals copied to `analysis_working/retrospective/`; outputs in `analysis_output/retrospective/`.

**R environment:** Knit with `/work_space/envs/transcriptomics2/bin/R` (limma on `/work_space` disk).

```{r setup}
library(dplyr)
library(readr)
if (requireNamespace("limma", quietly = TRUE)) library(limma)
```

```{r load-data}
if (file.exists("retrospective/Leuven_rlog_values.txt")) {
  base_dir       <- "."
  proteomics_dir <- "../proteomics"
} else if (file.exists("Transcriptomics/retrospective/Leuven_rlog_values.txt")) {
  base_dir       <- "Transcriptomics"
  proteomics_dir <- "proteomics"
} else {
  stop("Knit from Transcriptomics/ or its parent directory.")
}

source(file.path(base_dir, "analysis_helpers.R"))
dirs <- setup_analysis_dirs(base_dir, "retrospective")
output_dir <- dirs$output_dir

original_files <- c(
  expression       = file.path(base_dir, "retrospective/Leuven_rlog_values.txt"),
  survival w/ risk = file.path(proteomics_dir, "survival_df_with_risk.rds")
)
missing <- names(original_files)[!file.exists(original_files)]
if (length(missing) > 0) {
  stop(
    "Missing required original file(s):\n",
    paste0("  - ", missing, ": ", original_files[missing], collapse = "\n")
  )
}

working <- ensure_working_copies(original_files, dirs$work_dir)
survival_df <- readRDS(working["survival w/ risk"])
```

```{r build-matrix-and-meta}
expr_mat <- prepare_expr_genes_x_samples(
  read_tsv(working["expression"], show_col_types = FALSE),
  "Geneid"
)
expr_mat <- filter_impute_expr(expr_mat)
cat("Genes x samples:", nrow(expr_mat), "x", ncol(expr_mat), "\n")

sample_ids <- colnames(expr_mat)
sample_meta <- data.frame(
  sample_id  = sample_ids,
  patient_id = sample_ids,
  stringsAsFactors = FALSE
) %>%
  left_join(
    survival_df %>% select(patient_id, risk_grp, rsf_risk),
    by = "patient_id"
  )

cat("Samples with Low/High risk:\n")
print(table(sample_meta$risk_grp, useNA = "always"))
```

```{r limma-de}
limma_out <- run_limma_high_vs_low(
  expr_mat     = expr_mat,
  sample_meta  = sample_meta,
  output_dir   = output_dir,
  file_prefix  = "Retrospective"
)

cat("\nTop 5 up (High > Low):\n")
print(limma_out$top %>% filter(direction == "up"))

cat("\nTop 5 down (High < Low):\n")
print(limma_out$top %>% filter(direction == "down"))

sig <- limma_out$results %>% filter(adj.P.Val < 0.05)
cat("\nGenes FDR < 0.05:", nrow(sig), "\n")
if (nrow(sig) > 0) print(head(sig %>% select(gene, logFC, P.Value, adj.P.Val), 10))
```

```{r limma-figures}
export_limma_figures(
  results     = limma_out$results,
  expr_mat    = expr_mat,
  sample_meta = sample_meta,
  output_dir  = output_dir,
  file_prefix = "Retrospective",
  cohort_title = "Retrospective Transcriptomics"
)
```
