---
title: "Taxonomy limma DE by DFS risk group"
output: html_notebook
---

Gene-level differential expression (High vs Low DFS risk) for **Taxonomy RNA-Seq** using limma — same pipeline as Stage 2, Retrospective, and Colossus.

**Expression:** `Taxonomy_calls/Taxonomy_manuela_with_rna_rlog.txt` (samples × genes, rlog).

**Sample mapping:** `Manuela_and_Belfast_RNA_classifications_CMS_CRIS.txt` (`Patient` → `Code` / G `patient_id`).

**Risk groups:** `risk_grp` from `proteomics/survival_df_with_risk.rds`.

**Outputs:** `analysis_output/taxonomy/Taxonomy_limma_high_vs_low_all_genes.csv`

**R environment:** `/work_space/envs/transcriptomics2/bin/R`

```{r setup}
library(dplyr)
library(readr)
if (requireNamespace("limma", quietly = TRUE)) library(limma)
```

```{r load-data}
taxonomy_rlog_rel <- file.path(
  "Taxonomy_calls", "Taxonomy_manuela_with_rna_rlog.txt"
)
taxonomy_map_rel <- file.path(
  "Taxonomy_calls", "Manuela_and_Belfast_RNA_classifications_CMS_CRIS.txt"
)

if (file.exists(taxonomy_rlog_rel)) {
  base_dir       <- "."
  proteomics_dir <- "../proteomics"
} else if (file.exists(file.path("Transcriptomics", taxonomy_rlog_rel))) {
  base_dir       <- "Transcriptomics"
  proteomics_dir <- "proteomics"
} else {
  stop("Knit from Transcriptomics/ or its parent directory.")
}

source(file.path(base_dir, "analysis_helpers.R"))
dirs <- setup_analysis_dirs(base_dir, "taxonomy")
output_dir <- dirs$output_dir

original_files <- c(
  expression       = file.path(base_dir, taxonomy_rlog_rel),
  taxonomy ID map  = file.path(base_dir, taxonomy_map_rel),
  survival w/ risk = file.path(proteomics_dir, "survival_df_with_risk.rds")
)
missing <- names(original_files)[!file.exists(original_files)]
if (length(missing) > 0) {
  stop(
    "Missing required original file(s):\n",
    paste0("  - ", missing, ": ", original_files[missing], collapse = "\n")
  )
}

working <- ensure_working_copies(original_files, dirs$work_dir)
```

```{r build-matrix-and-meta}
cat("Loading Taxonomy rlog (large file)...\n")
expr_mat <- load_taxonomy_expression_matrix(working["expression"])
cat("Genes x samples:", nrow(expr_mat), "x", ncol(expr_mat), "\n")

sample_meta <- build_taxonomy_sample_meta(
  working["taxonomy ID map"],
  working["survival w/ risk"]
) %>%
  filter(sample_id %in% colnames(expr_mat))

cat("Samples with Low/High risk:\n")
print(table(sample_meta$risk_grp[match(colnames(expr_mat), sample_meta$sample_id)], useNA = "always"))
```

```{r limma-de}
limma_out <- run_limma_high_vs_low(
  expr_mat     = expr_mat,
  sample_meta  = sample_meta,
  output_dir   = output_dir,
  file_prefix  = "Taxonomy"
)

cat("\nTop 5 up (High > Low):\n")
print(limma_out$top %>% filter(direction == "up"))

cat("\nTop 5 down (High < Low):\n")
print(limma_out$top %>% filter(direction == "down"))

sig <- limma_out$results %>% filter(adj.P.Val < 0.05)
cat("\nGenes FDR < 0.05:", nrow(sig), "\n")
if (nrow(sig) > 0) print(head(sig %>% select(gene, logFC, P.Value, adj.P.Val), 10))
```

```{r limma-figures}
export_limma_figures(
  results      = limma_out$results,
  expr_mat       = expr_mat,
  sample_meta    = sample_meta,
  output_dir     = output_dir,
  file_prefix    = "Taxonomy",
  cohort_title   = "Taxonomy RNA-Seq"
)
```
