---
title: "xCell combined enrichment after per-cell-type Anderson-Darling tests"
output: html_notebook
---

Tests whether each xCell2 cell type's enrichment score **distribution** is similar **across the four cohorts** (Anderson-Darling k-sample test per cell type).

This is **not** stratified by High/Low risk — it only asks whether Stage2, Retrospective, Colossus, and Taxonomy have comparable score distributions for each cell feature.

If any cell types pass (AD p ≥ 0.05), those are pooled and High vs Low Wilcoxon, volcano, and top-5 tables are rerun on the passing subset only.

**Outputs:** `analysis_output/combined/`

```{r setup}
library(dplyr)
library(readr)
source("analysis_helpers.R")
source("xcell_combine_helpers.R")
```

```{r ad-test}
if (dir.exists("analysis_output/combined")) {
  base_dir <- "."
} else if (dir.exists("Transcriptomics/analysis_output/combined")) {
  base_dir <- "Transcriptomics"
} else {
  base_dir <- "."
}
output_dir <- file.path(base_dir, "analysis_output", "combined")

out <- run_xcell_ad_and_combined(
  output_dir = output_dir,
  ad_alpha = 0.05
)

print(out$ad_out$results %>% arrange(ad_p) %>% head(15))
cat("\nPassing:", length(out$passing_cell_types), "cell types\n")
```

```{r combined-results}
if (out$can_combine && !is.null(out$diff_combined)) {
  cat("\nCombined Wilcoxon (AD-passing cell types) — top hits:\n")
  print(
    out$diff_combined %>%
      arrange(pval) %>%
      select(cell_type, log2FC, pval) %>%
      head(10)
  )
} else {
  cat("\nNo cell types passed AD — review xcell_ad_by_cell_type.csv\n")
}
```

```{r density-plots}
density_dir <- file.path(output_dir, "density_plots")
export_xcell_cohort_density_plots(
  out$scores_long,
  output_dir = density_dir,
  n_cell_types = 12L,
  z_score_within_cell_type = FALSE
)
export_xcell_cohort_density_plots(
  out$scores_long,
  output_dir = density_dir,
  n_cell_types = 12L,
  z_score_within_cell_type = TRUE
)
```

```{r quantile-combined}
qn_out <- run_xcell_quantile_combined_analysis(output_dir = output_dir)
print(
  qn_out$diff_combined %>%
    arrange(pval) %>%
    select(cell_type, log2FC, pval) %>%
    head(10)
)
```
