---
title: "Adjuvant-treated patients: RSF metacluster risk vs DFS relapse"
output: html_notebook
---

Among **adjuvant-treated** patients, test whether **RSF-derived metacluster risk** and underlying **metacluster composition** differ between those who **relapse** (`dfs_event = 1`) and those who **do not** (`dfs_event = 0`).

**Risk definition (pre-computed):** `survival_df_with_risk.rds` from `RCSI_Taxonomy_Combined_prognosis_scores.Rmd`

- **`rsf_risk`:** predicted risk from random survival forest (`randomForestSRC`) fit on metacluster probability columns.
- **`risk_grp`:** Low / High from optimal DFS cutpoint on `rsf_risk` (`survminer::surv_cutpoint`, min group size 15%).

**Inclusion:** `has_adjuvant_chemotherapy == "yes"`, valid `dfs_months > 0`, non-missing `dfs_event`, and non-missing RSF risk.

**Outputs:** `results/adjuvant_metacluster_relapse/` (CSVs + PNG/PDF figures).

Knit from the `proteomics/` directory.

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(readr)
library(stringr)
library(ggplot2)
library(ggpubr)
library(survival)
library(broom)

if (requireNamespace("pheatmap", quietly = TRUE)) library(pheatmap)

output_dir <- "results/adjuvant_metacluster_relapse"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

save_csv <- function(x, name) {
  path <- file.path(output_dir, name)
  write_csv(x, path)
  message("Saved: ", path)
}

save_plot <- function(p, name, width = 7, height = 6) {
  png_path <- file.path(output_dir, sub("\\.pdf$", ".png", name))
  pdf_path <- file.path(output_dir, sub("\\.png$", ".pdf", name))
  ggsave(png_path, p, width = width, height = height, dpi = 300, bg = "white")
  ggsave(pdf_path, p, width = width, height = height, bg = "white")
  message("Saved: ", png_path)
}
```

```{r load-data}
merged_df <- readRDS("RCSI_Taxonomy_clin_prot_merged_df.rds")
survival_df <- readRDS("survival_df_with_risk.rds")

stopifnot(all(c("patient_id", "rsf_risk", "risk_grp") %in% names(survival_df)))

leiden_cols_all <- grep(
  "abundance_metacluster_leiden",
  colnames(merged_df),
  value = TRUE
)

cat("Merged cohort:", nrow(merged_df), "patients\n")
cat("RSF risk table:", nrow(survival_df), "patients\n")
cat("Metacluster columns:", length(leiden_cols_all), "\n")
print(table(survival_df$risk_grp, useNA = "ifany"))
```

```{r define-cohort}
adj_df <- merged_df %>%
  filter(
    has_adjuvant_chemotherapy == "yes",
    !is.na(dfs_months),
    dfs_months > 0,
    !is.na(dfs_event)
  ) %>%
  left_join(
    survival_df %>%
      select(patient_id, rsf_risk, risk_grp, dfs_months, dfs_event),
    by = "patient_id",
    suffix = c("", "_rsf")
  ) %>%
  filter(!is.na(rsf_risk), !is.na(risk_grp)) %>%
  mutate(
    relapse_status = factor(
      dfs_event,
      levels = c(0, 1),
      labels = c("No relapse", "Relapse")
    ),
    risk_grp = factor(risk_grp, levels = c("Low", "High"))
  )

zero_frac <- colMeans(adj_df[, leiden_cols_all] == 0, na.rm = TRUE)
leiden_cols <- leiden_cols_all[zero_frac < 1]

cat("\nAdjuvant-treated patients with RSF risk:", nrow(adj_df), "\n")
print(table(adj_df$relapse_status, adj_df$risk_grp))

cat("\nRelapse rate by RSF risk group:\n")
relapse_rate_table <- adj_df %>%
  group_by(risk_grp) %>%
  summarise(
    n = n(),
    n_relapse = sum(dfs_event == 1),
    relapse_rate = mean(dfs_event == 1),
    .groups = "drop"
  )
print(relapse_rate_table)
save_csv(relapse_rate_table, "adjuvant_rsf_risk_group_relapse_rates.csv")
```

```{r rsf-score-tests}
wt_risk <- wilcox.test(
  adj_df$rsf_risk[adj_df$dfs_event == 1],
  adj_df$rsf_risk[adj_df$dfs_event == 0],
  exact = FALSE
)

cat("\n=== Wilcoxon: rsf_risk (Relapse vs No relapse) ===\n")
cat("  Median rsf_risk (Relapse):   ", median(adj_df$rsf_risk[adj_df$dfs_event == 1]), "\n")
cat("  Median rsf_risk (No relapse):", median(adj_df$rsf_risk[adj_df$dfs_event == 0]), "\n")
cat("  p-value:", format(wt_risk$p.value, digits = 3), "\n")

glm_risk <- glm(dfs_event ~ rsf_risk, data = adj_df, family = binomial())
glm_tidy <- tidy(glm_risk, conf.int = TRUE, exponentiate = TRUE)

cat("\n=== Logistic regression: dfs_event ~ rsf_risk ===\n")
print(glm_tidy)
save_csv(glm_tidy, "adjuvant_rsf_risk_logistic_regression.csv")

tbl_risk <- table(
  Relapse = adj_df$relapse_status,
  RSF_risk = adj_df$risk_grp
)

cat("\n=== RSF risk group vs relapse ===\n")
print(tbl_risk)

chi_risk <- chisq.test(tbl_risk)
fisher_risk <- fisher.test(tbl_risk)

relapse_table <- adj_df %>%
  count(risk_grp, relapse_status) %>%
  pivot_wider(names_from = relapse_status, values_from = n, values_fill = 0) %>%
  mutate(
    n_total = `No relapse` + Relapse,
    relapse_rate_pct = round(100 * Relapse / n_total, 1)
  )

save_csv(relapse_table, "adjuvant_rsf_risk_group_relapse_table.csv")

results_summary <- tibble(
  test = c(
    "Wilcoxon rsf_risk",
    "Logistic rsf_risk OR",
    "Chi-square risk_grp",
    "Fisher risk_grp OR"
  ),
  statistic = c(
    NA_real_,
    glm_tidy$estimate[glm_tidy$term == "rsf_risk"],
    chi_risk$statistic,
    fisher_risk$estimate
  ),
  p_value = c(
    wt_risk$p.value,
    glm_tidy$p.value[glm_tidy$term == "rsf_risk"],
    chi_risk$p.value,
    fisher_risk$p.value
  )
)

save_csv(results_summary, "adjuvant_rsf_risk_relapse_tests.csv")
print(results_summary)
```

```{r rsf-score-boxplot}
relapse_colours <- c("No relapse" = "#1B4332", "Relapse" = "#B2182B")

p_score <- ggplot(adj_df, aes(x = relapse_status, y = rsf_risk, fill = relapse_status)) +
  geom_boxplot(alpha = 0.85, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.35, size = 1.2) +
  scale_fill_manual(values = relapse_colours) +
  stat_compare_means(method = "wilcox.test", label = "p.format", size = 4) +
  labs(
    title = "RSF metacluster risk score by relapse status",
    subtitle = paste0(
      "Adjuvant-treated only | n = ", nrow(adj_df),
      " (No relapse ", sum(adj_df$dfs_event == 0),
      ", Relapse ", sum(adj_df$dfs_event == 1), ")"
    ),
    x = NULL,
    y = "RSF predicted risk (rsf_risk)"
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "none")

print(p_score)
save_plot(p_score, "adjuvant_rsf_risk_by_relapse_boxplot.png")
```

```{r risk-group-barplot}
relapse_by_group <- adj_df %>%
  count(risk_grp, relapse_status) %>%
  group_by(risk_grp) %>%
  mutate(pct = n / sum(n) * 100)

p_bar <- ggplot(relapse_by_group, aes(x = risk_grp, y = pct, fill = relapse_status)) +
  geom_col(position = "stack", colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = relapse_colours, name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Relapse proportion by RSF risk group",
    subtitle = "Adjuvant-treated patients only",
    x = "RSF risk group",
    y = "Percentage within group"
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")

print(p_bar)
save_plot(p_bar, "adjuvant_rsf_risk_group_relapse_barplot.png")
```

```{r km-adjuvant-subset}
km_adj <- survfit(Surv(dfs_months, dfs_event) ~ risk_grp, data = adj_df)

if (requireNamespace("survminer", quietly = TRUE)) {
  p_km <- survminer::ggsurvplot(
    km_adj,
    data = adj_df,
    pval = TRUE,
    risk.table = TRUE,
    palette = c("#004B23", "#800000"),
    legend.labs = c("Low", "High"),
    legend.title = "RSF risk",
    xlab = "Time (months)",
    ylab = "DFS probability",
    title = "DFS in adjuvant-treated patients by RSF risk group",
    ggtheme = theme_minimal()
  )
  print(p_km)

  km_png <- file.path(output_dir, "adjuvant_rsf_risk_km_curve.png")
  km_pdf <- file.path(output_dir, "adjuvant_rsf_risk_km_curve.pdf")
  png(km_png, width = 9, height = 8, units = "in", res = 300)
  print(p_km)
  dev.off()
  pdf(km_pdf, width = 9, height = 8)
  print(p_km)
  dev.off()
  message("Saved: ", km_png)
}

logrank_adj <- survdiff(Surv(dfs_months, dfs_event) ~ risk_grp, data = adj_df)
cat("\nLog-rank (adjuvant subset):\n")
print(logrank_adj)
```

```{r mc-per-cluster-wilcoxon}
mc_prob_stats <- map_dfr(leiden_cols, function(col) {
  pr <- adj_df[[col]][adj_df$dfs_event == 1]
  pn <- adj_df[[col]][adj_df$dfs_event == 0]
  pr <- pr[!is.na(pr)]
  pn <- pn[!is.na(pn)]

  if (length(pr) < 2 || length(pn) < 2) {
    return(tibble(column = col, metacluster = NA_character_, p_value = NA_real_))
  }

  wt <- wilcox.test(pr, pn, exact = FALSE)

  tibble(
    column = col,
    metacluster = gsub("abundance_metacluster_leiden_1.0_", "", col),
    mean_relapse = mean(pr),
    mean_no_relapse = mean(pn),
    median_relapse = median(pr),
    median_no_relapse = median(pn),
    delta_mean = mean(pr) - mean(pn),
    cor_rsf = cor(adj_df[[col]], adj_df$rsf_risk, use = "complete.obs"),
    p_value = wt$p.value
  )
}) %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH")) %>%
  arrange(p_value)

save_csv(mc_prob_stats, "adjuvant_mc_relapse_wilcoxon.csv")

cat("\nMetaclusters tested:", sum(is.finite(mc_prob_stats$p_value)), "\n")
cat("Raw p < 0.05:", sum(mc_prob_stats$p_value < 0.05, na.rm = TRUE), "\n")
cat("FDR < 0.10:", sum(mc_prob_stats$adj_p_value < 0.10, na.rm = TRUE), "\n")
cat("\nTop 10 by raw p-value:\n")
print(head(mc_prob_stats, 10))
```

```{r mc-composite-tests}
composite_results <- tibble(test = character(), statistic = double(), p_value = double())

if (file.exists("mc_event_both.rds") && file.exists("mc_noevent_both.rds")) {
  bad_cols <- intersect(readRDS("mc_event_both.rds"), leiden_cols)
  good_cols <- intersect(readRDS("mc_noevent_both.rds"), leiden_cols)

  adj_df <- adj_df %>%
    mutate(
      bad_mc_sum = rowSums(across(all_of(bad_cols)), na.rm = TRUE),
      good_mc_sum = rowSums(across(all_of(good_cols)), na.rm = TRUE),
      bad_minus_good = bad_mc_sum - good_mc_sum
    )

  wt_bad <- wilcox.test(
    adj_df$bad_mc_sum[adj_df$dfs_event == 1],
    adj_df$bad_mc_sum[adj_df$dfs_event == 0],
    exact = FALSE
  )
  wt_good <- wilcox.test(
    adj_df$good_mc_sum[adj_df$dfs_event == 1],
    adj_df$good_mc_sum[adj_df$dfs_event == 0],
    exact = FALSE
  )
  wt_comp <- wilcox.test(
    adj_df$bad_minus_good[adj_df$dfs_event == 1],
    adj_df$bad_minus_good[adj_df$dfs_event == 0],
    exact = FALSE
  )

  composite_results <- tibble(
    test = c(
      "Event-enriched MC sum",
      "No-event-enriched MC sum",
      "Bad minus good composite"
    ),
    statistic = c(
      median(adj_df$bad_mc_sum[adj_df$dfs_event == 1]) -
        median(adj_df$bad_mc_sum[adj_df$dfs_event == 0]),
      median(adj_df$good_mc_sum[adj_df$dfs_event == 1]) -
        median(adj_df$good_mc_sum[adj_df$dfs_event == 0]),
      median(adj_df$bad_minus_good[adj_df$dfs_event == 1]) -
        median(adj_df$bad_minus_good[adj_df$dfs_event == 0])
    ),
    p_value = c(wt_bad$p.value, wt_good$p.value, wt_comp$p.value)
  )

  save_csv(composite_results, "adjuvant_mc_composite_relapse_tests.csv")
  print(composite_results)
}
```

```{r mc-global-pca-test}
prob_mat <- as.matrix(adj_df[, leiden_cols])
pca <- prcomp(prob_mat, center = TRUE, scale. = FALSE)

n_pc <- min(5, ncol(prob_mat))
pc_df <- as.data.frame(pca$x[, seq_len(n_pc), drop = FALSE])
pc_df$relapse <- adj_df$dfs_event

pc_formula <- as.formula(
  paste("relapse ~", paste(colnames(pc_df)[seq_len(n_pc)], collapse = " + "))
)
global_glm <- glm(pc_formula, data = pc_df, family = binomial())
pc_anova <- as.data.frame(anova(global_glm, test = "Chisq")) %>%
  rownames_to_column("term")

save_csv(pc_anova, "adjuvant_mc_pca_logistic_anova.csv")

cat("\n=== Global logistic regression on top", n_pc, "PCs ===\n")
print(pc_anova)

pc_plot_df <- pc_df %>%
  mutate(relapse_status = adj_df$relapse_status)

p_pca <- ggplot(pc_plot_df, aes(x = PC1, y = PC2, colour = relapse_status)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse(linewidth = 0.7) +
  scale_colour_manual(values = relapse_colours, name = NULL) +
  labs(
    title = "Metacluster composition (PCA)",
    subtitle = "Adjuvant-treated patients coloured by relapse status",
    x = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "% var)"),
    y = paste0("PC2 (", round(100 * summary(pca)$importance[2, 2], 1), "% var)")
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p_pca)
save_plot(p_pca, "adjuvant_mc_pca_relapse.png", width = 7, height = 6)
```

```{r mc-volcano-plot}
volcano_df <- mc_prob_stats %>%
  filter(is.finite(delta_mean), is.finite(p_value), p_value > 0)

p_mc_volcano <- ggplot(
  volcano_df,
  aes(x = delta_mean, y = -log10(p_value), colour = adj_p_value < 0.10)
) +
  geom_point(alpha = 0.75, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(
    values = c("FALSE" = "grey60", "TRUE" = "#B2182B"),
    name = "FDR < 0.10"
  ) +
  labs(
    title = "Metacluster probability shift by relapse status",
    subtitle = "Adjuvant-treated patients | Wilcoxon per metacluster",
    x = "Mean difference (Relapse - No relapse)",
    y = expression("-log"[10] * "(p-value)")
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")

if (requireNamespace("ggrepel", quietly = TRUE)) {
  label_df <- volcano_df %>% arrange(p_value) %>% slice_head(n = 12)
  p_mc_volcano <- p_mc_volcano +
    ggrepel::geom_text_repel(
      data = label_df,
      aes(label = metacluster),
      size = 3,
      max.overlaps = Inf,
      show.legend = FALSE,
      colour = "black"
    )
}

print(p_mc_volcano)
save_plot(p_mc_volcano, "adjuvant_mc_relapse_volcano.png", width = 9, height = 6)
```

```{r mc-heatmap-top}
top_n <- 20
top_mc <- mc_prob_stats %>%
  filter(is.finite(p_value)) %>%
  arrange(p_value) %>%
  slice_head(n = top_n) %>%
  pull(column)

if (length(top_mc) >= 2 && requireNamespace("pheatmap", quietly = TRUE)) {
  hm_mat <- adj_df %>%
    group_by(relapse_status) %>%
    summarise(across(all_of(top_mc), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
    column_to_rownames("relapse_status") %>%
    t()

  rownames(hm_mat) <- gsub("abundance_metacluster_leiden_1.0_", "", rownames(hm_mat))
  hm_mat <- hm_mat[apply(hm_mat, 1, sd) > 0, , drop = FALSE]

  draw_hm <- function() {
    pheatmap(
      hm_mat,
      scale = "row",
      cluster_cols = FALSE,
      cluster_rows = TRUE,
      border_color = NA,
      color = colorRampPalette(c("#1B4332", "white", "#B2182B"))(100),
      main = paste0("Top ", nrow(hm_mat), " metaclusters by Wilcoxon p-value"),
      fontsize_row = 9
    )
  }

  hm_png <- file.path(output_dir, "adjuvant_mc_relapse_heatmap.png")
  hm_pdf <- file.path(output_dir, "adjuvant_mc_relapse_heatmap.pdf")
  png(hm_png, width = 8, height = 9, units = "in", res = 300)
  draw_hm()
  dev.off()
  pdf(hm_pdf, width = 8, height = 9)
  draw_hm()
  dev.off()
  message("Saved: ", hm_png)
}
```

```{r mc-boxplots-top}
top_hits <- mc_prob_stats %>%
  filter(is.finite(p_value)) %>%
  arrange(p_value) %>%
  slice_head(n = 6) %>%
  pull(column)

if (length(top_hits) > 0) {
  long_box <- adj_df %>%
    select(relapse_status, all_of(top_hits)) %>%
    pivot_longer(-relapse_status, names_to = "column", values_to = "probability") %>%
    left_join(
      mc_prob_stats %>% select(column, metacluster, p_value, cor_rsf),
      by = "column"
    ) %>%
    mutate(
      label = paste0(metacluster, "\np=", signif(p_value, 2))
    )

  p_mc_box <- ggplot(long_box, aes(x = relapse_status, y = probability, fill = relapse_status)) +
    geom_boxplot(outlier.shape = 21, alpha = 0.85) +
    geom_jitter(width = 0.12, alpha = 0.35, size = 1) +
    facet_wrap(~ label, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = relapse_colours) +
    labs(
      title = "Top metaclusters: probability by relapse status",
      subtitle = "Adjuvant-treated patients only",
      x = NULL,
      y = "Metacluster probability"
    ) +
    theme_bw(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "none",
      strip.text = element_text(size = 9)
    )

  print(p_mc_box)
  save_plot(p_mc_box, "adjuvant_mc_relapse_top_boxplots.png", width = 10, height = 7)
}
```

```{r mc-composite-boxplot}
if ("bad_minus_good" %in% names(adj_df)) {
  p_comp <- ggplot(adj_df, aes(x = relapse_status, y = bad_minus_good, fill = relapse_status)) +
    geom_boxplot(alpha = 0.85, outlier.shape = 21) +
    geom_jitter(width = 0.12, alpha = 0.35, size = 1.2) +
    scale_fill_manual(values = relapse_colours) +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 4) +
    labs(
      title = "Event-enriched minus no-event metacluster composite",
      subtitle = "Adjuvant-treated patients by relapse status",
      x = NULL,
      y = "Bad MC sum - Good MC sum"
    ) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(face = "bold"), legend.position = "none")

  print(p_comp)
  save_plot(p_comp, "adjuvant_mc_bad_minus_good_boxplot.png")
}
```

```{r summary}
cat("\n=== Summary ===\n")
cat("Output directory:", output_dir, "\n")
cat("Adjuvant patients analysed:", nrow(adj_df), "\n")
cat("  Low risk: ", sum(adj_df$risk_grp == "Low"), "\n")
cat("  High risk:", sum(adj_df$risk_grp == "High"), "\n")
cat("  Relapse:  ", sum(adj_df$dfs_event == 1), "\n")
cat("\nRSF risk tests:\n")
print(results_summary)
if (exists("composite_results") && nrow(composite_results) > 0) {
  cat("\nComposite metacluster tests:\n")
  print(composite_results)
}
cat("\nFiles written to ", output_dir, ":\n", sep = "")
print(list.files(output_dir))
```
